---
title: "Example R Code for SoTL projects"
author: "Ashley Yue Mao"
date: "`r Sys.Date()`"
output: html_document
---

# 0. Install and load required packages

Packages you might use: - **tidyverse**: a collection of packages that you're likely to use in everyday data analyses.

In order to use a package, you need to install it to your computer (once) and load it every time when you start a new R session. As an example, here is how you can install and load a package:

```{r install a package}
install.packages("tidyverse")
```

```{r load a package}
library(tidyverse)
```

# 1. Read in your data

Download survey data as .csv files from [Qualtrics](https://it.ubc.ca/services/teaching-learning-tools/survey-tool) and then you can use `read_csv` functions to read in your data.

In this document, I will use three simulated data file: - a file that contain a identifier for each student, and their grades before the intervention - a file that contain a identifier for each student, and their grades after the intervention - a file that contain a identifier for each strudent, and their gender and group (e.g., experimental, control)

```{r read in data}
pre <- read_csv("data/pre_grades.csv")
post <- read_csv("data/post_grades.csv")
info <- read_csv("data/info.csv")
```

As for how to import other types of data, such as SPSS, SAS, JASON, check the [datacamp](https://www.datacamp.com/tutorial/r-data-import-tutorial).

Let's have a look on the data:

```{r data overview}
str(pre)
str(post)
str(info)

View(pre)
View(post)
View(info)
```

# 2. Combine the three dataset into one

A common task then is to combine the three dataset into one for further analysis.

Step1: combine pre grades and post grades into one dataset
-   rename column names in pre and post grades data to be consistent
-   add a new column to keep record which time do the grades belong to in the new dataset

```{r combine pre and post grades in long format}
pre <- pre |>
  rename("identifier" = "Identifier",
         "grades" = "Scores") |>
  add_column(time = "pre")

post <- post |>
  add_column(time = "post")

grades_long <- pre |>
  bind_rows(post)
```

```{r combine pre and post grades in wide
format}
pre <- pre |>
  rename("identifier" = "Identifier",
         "grades" = "Scores") |>
  add_column(time = "pre")

post <- post |>
  add_column(time = "post")

grades_long <- pre |>
  bind_rows(post)
```




# 3. Clean the data

# 4. Descriptive stats

# 5. Visualization

# 6. Statistical analysis

# 7. Reporting
